---
title: "Figure 3 panel — Brain expression heatmap (top SFARI-1 genes)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
params:
  regenerate:        false
  panel_id:          "Figure_3_brain_heatmap"
  hasura_url:        "http://localhost:8789/v1/graphql"
  env_file:          ""
  cache_db:          ""
  intermediate_dir:  ""
  out_main:          ""
  out_supp_tables:   ""
  n_top_genes:       20
  organ_id_brain:    8
  panel_tag:         "A"
  width:             8
  height:            10
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

root <- Sys.getenv("ASD_PAPER_ROOT")
if (!nzchar(root)) stop("ASD_PAPER_ROOT is unset -- see figures/README.md")
source(file.path(root, "figures", "_paths.R"))
suppressPackageStartupMessages({
  library(httr); library(jsonlite); library(dplyr); library(tidyr)
  library(ggplot2); library(scales); library(tibble)
})
source(asd_code("ver1", "_helpers.R"))

panel_id        <- params$panel_id
intermediate_dir<- asd_param_path(params$intermediate_dir, asd_intermediate_dir())
out_main        <- asd_param_path(params$out_main,         asd_panels_dir())
out_supp_tables <- asd_param_path(params$out_supp_tables,  asd_supp_tables_dir("panels"))
dir.create(intermediate_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(out_main,         showWarnings = FALSE, recursive = TRUE)
dir.create(out_supp_tables,  showWarnings = FALSE, recursive = TRUE)

expr_rds       <- file.path(intermediate_dir, "brain_expr.rds")
top_genes_rds  <- file.path(intermediate_dir, "top_genes.rds")
ggplot_rds     <- file.path(intermediate_dir, paste0(panel_id, "_ggplot.rds"))
source_csv     <- file.path(intermediate_dir, paste0("Source_data_", panel_id, ".csv"))
```

# Plan

Adapted from `ASD_code_try_ver4_130326_Nave_dataset.rmd` "Figure 2B" — brain
expression heatmap of the top SFARI-score-1 genes by variant count.  Two
versions exist in the v4 code; the collaborators asked for the **clean** one (rows
labelled "GENE" only, no SFARI annotation).  Tag = `A` of Figure 3.

Inputs:

- Top genes: `params$n_top_genes` SFARI-1 gene_symbols ranked by variant
  count in `df_cds`.
- Expression: `gene_expressions` from the DB filtered to `organ_id = 8`
  (Brain), pulled via `gql_read()` and pivoted to a gene × tissue matrix.

Tissues shown (column order — matches v4):

1. Cortex
2. Frontal Cortex (BA9)
3. Anterior cingulate cortex (BA24)
4. Cerebellar Hemisphere
5. Cerebellum

Heatmap fill: `log10(TPM + 1)`, white → blue gradient.

# Read-only DB wrapper

```{r gql_read}
.load_hasura_secret <- function(env_file) {
  if (Sys.getenv("HASURA_ADMIN_SECRET") != "") return(invisible(NULL))
  if (!file.exists(env_file)) return(invisible(NULL))
  for (l in readLines(env_file, warn = FALSE)) {
    l <- trimws(l)
    if (l == "" || startsWith(l, "#")) next
    m <- regmatches(l, regexec("^HASURA_ADMIN_SECRET=\"?([^\"]*)\"?$", l))[[1]]
    if (length(m) == 2) { Sys.setenv(HASURA_ADMIN_SECRET = m[2]); break }
  }
}
.load_hasura_secret(asd_param_path(params$env_file, asd_env_file()))
HASURA_URL    <- params$hasura_url
HASURA_SECRET <- Sys.getenv("HASURA_ADMIN_SECRET")
stopifnot(nchar(HASURA_SECRET) > 0)

gql_read <- function(query, max_retries = 5) {
  q <- trimws(query)
  if (!grepl("^(query\\b|\\{)", q))
    stop("gql_read: query must start with 'query' or '{'.")
  for (tok in c("mutation ", "mutation{", "insert_", "update_", "delete_")) {
    if (grepl(tok, tolower(q), fixed = TRUE))
      stop("gql_read: forbidden token '", tok, "'.")
  }
  body <- toJSON(list(query = query), auto_unbox = TRUE)
  for (attempt in seq_len(max_retries)) {
    resp <- tryCatch(
      POST(HASURA_URL,
           add_headers(`x-hasura-admin-secret` = HASURA_SECRET),
           content_type_json(), body = body, timeout(180)),
      error = function(e) { cat("[gql_read] attempt", attempt, "POST error:",
                                conditionMessage(e), "\n"); NULL })
    if (is.null(resp))                       { Sys.sleep(0.5 * 2^(attempt-1)); next }
    if (status_code(resp) %in% c(502,503,504)) { Sys.sleep(0.5 * 2^(attempt-1)); next }
    parsed <- fromJSON(content(resp, as = "text", encoding = "UTF-8"), flatten = TRUE)
    if (!is.null(parsed$errors))
      stop("GraphQL error: ", toJSON(parsed$errors, auto_unbox = TRUE))
    return(parsed$data)
  }
  stop("gql_read: failed after ", max_retries, " attempts")
}
```

# Top genes (from cached df_cds)

```{r top_genes}
df_cds <- readRDS(asd_param_path(params$cache_db, asd_db_snapshot("df_cds.rds")))

top_genes_tbl <- df_cds %>%
  filter(sfari_gene_score == "1") %>%
  count(gene_symbol, name = "total_variants") %>%
  arrange(desc(total_variants)) %>%
  head(params$n_top_genes)

top_genes <- top_genes_tbl$gene_symbol
saveRDS(top_genes_tbl, top_genes_rds)
cat("top", params$n_top_genes, "SFARI-1 genes by variant count:\n")
print(top_genes_tbl)
```

# Pull (or reload) brain expression

```{r pull_expression}
have_cache <- file.exists(expr_rds)
if (!params$regenerate && have_cache) {
  cat("reusing cache:", expr_rds, "\n")
  df_expr <- readRDS(expr_rds)
} else {
  cat("pulling gene expressions for organ_id =", params$organ_id_brain, "...\n")
  # Build _in array literal for the genes we care about
  syms_lit <- paste0('"', top_genes, '"', collapse = ",")
  q <- sprintf('{
    genes(where: { symbol: { _in: [%s] } }) {
      symbol
      gene_expressions(where: { tissue: { organ_id: { _eq: %d } } }) {
        tissue { name }
        median_tpm
      }
    }
  }', syms_lit, params$organ_id_brain)

  rs <- gql_read(q)$genes
  # rs$gene_expressions is a list-column (data.frame per gene) when flatten=TRUE
  rows <- lapply(seq_len(nrow(rs)), function(i) {
    sym <- rs$symbol[i]
    ge  <- rs$gene_expressions[[i]]
    if (is.null(ge) || nrow(ge) == 0)
      return(data.frame(gene_symbol = sym, tissue = character(0),
                        median_tpm = numeric(0)))
    data.frame(gene_symbol = sym,
               tissue      = ge$tissue.name,
               median_tpm  = as.numeric(ge$median_tpm))
  })
  df_expr <- bind_rows(rows)

  saveRDS(df_expr, expr_rds)
  cat("pulled:", nrow(df_expr), "gene×tissue rows ->", expr_rds, "\n")
}
print(head(df_expr))
```

# Build matrix and plot

```{r build_panel, fig.width=params$width, fig.height=params$height}
tissue_order <- c("Cortex",
                  "Frontal Cortex (BA9)",
                  "Anterior cingulate cortex (BA24)",
                  "Cerebellar Hemisphere",
                  "Cerebellum")

mat <- df_expr %>%
  filter(tissue %in% tissue_order) %>%
  pivot_wider(names_from = tissue, values_from = median_tpm) %>%
  column_to_rownames("gene_symbol") %>%
  as.matrix()

# Order columns and rows
mat <- mat[, intersect(tissue_order, colnames(mat))]
# Row order: top of heatmap = highest variant count (matches v4's bar order)
row_order <- intersect(rev(top_genes), rownames(mat))
mat <- mat[row_order, ]

cat("Heatmap matrix:", nrow(mat), "genes x", ncol(mat), "tissues\n")
cat("TPM range:", round(min(mat, na.rm = TRUE), 2), "—",
    round(max(mat, na.rm = TRUE), 2), "\n")

# Two-line tissue labels for the BA-suffixed columns to save horizontal space
display_tissue <- gsub(" \\(BA", "\n(BA", colnames(mat))
colnames(mat)  <- display_tissue

mat_long <- as.data.frame(mat) %>%
  rownames_to_column("gene_symbol") %>%
  pivot_longer(-gene_symbol, names_to = "tissue", values_to = "TPM") %>%
  mutate(gene_symbol = factor(gene_symbol, levels = row_order),
         tissue      = factor(tissue, levels = display_tissue))

p <- ggplot(mat_long, aes(x = tissue, y = gene_symbol,
                          fill = log10(TPM + 1))) +
  geom_tile(colour = "black", linewidth = 0.5) +
  scale_fill_gradientn(
    colours = c("white", "#deebf7", "#9ecae1", "#3182bd"),
    values  = scales::rescale(c(0, 0.33, 0.67, 1.0)),
    limits  = c(0, 2),
    oob     = scales::squish,
    na.value = "grey90",
    name    = "log10(TPM + 1)",
    breaks  = c(0, 0.5, 1, 1.5, 2),
    labels  = c("0", "0.5", "1", "1.5", "2")
  ) +
  labs(x = NULL, y = NULL, tag = params$panel_tag) +
  theme_publication(base_size = 12, x_angle = 45, y_bold = TRUE) +
  theme(
    axis.text.x       = element_text(face = "bold", angle = 45,
                                     hjust = 1, size = 11,
                                     family = "Nimbus Sans"),
    axis.text.y       = element_text(face = "bold", size = 11,
                                     family = "Nimbus Sans"),
    panel.grid        = element_blank(),
    legend.position   = "right",
    plot.tag          = element_text(face = "bold", size = 16,
                                     family = "Nimbus Sans"),
    plot.tag.position = c(0, 1)
  )
# print(p) below would fail under Rscript's null device on systems where
# the bold/italic Nimbus Sans variants trigger a CID-font lookup; saving
# via cairo_pdf works fine, so we skip the on-screen render.
if (interactive()) print(p) else invisible(p)
```

# Save

```{r save}
saveRDS(p, ggplot_rds)
out_pdf <- file.path(out_main, paste0(panel_id, ".pdf"))
out_png <- file.path(intermediate_dir, paste0(panel_id, ".png"))
ggsave(out_pdf, p, width = params$width, height = params$height,
       device = cairo_pdf, bg = "transparent")
ggsave(out_png, p, width = params$width, height = params$height,
       dpi = 150, bg = "white")
write.csv(mat_long, source_csv, row.names = FALSE)
cat("saved:\n  ", out_pdf, "\n  ", ggplot_rds, "\n  ", out_png, "\n  ",
    source_csv, "\n")
```
