---
title: "ASD ADAR Paper — 00 Prep Data (DB pull + cache)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
params:
  regenerate:        false                                   # default: read cached snapshot
  cache_dir:        ""      # default: <out>/Intermediate_tables/00_db_snapshot
  hasura_url:       "http://localhost:8789/v1/graphql"       # check stack on alu17 (the populated one; check2 is a partial test DB)
  env_file:         ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

root <- Sys.getenv("ASD_PAPER_ROOT")
if (!nzchar(root)) stop("ASD_PAPER_ROOT is unset -- see figures/README.md")
source(file.path(root, "figures", "_paths.R"))

suppressPackageStartupMessages({
  library(httr)
  library(jsonlite)
  library(dplyr)
  library(tidyr)
  library(stringr)
  library(digest)
})

# --- Where the cache lives ---
cache_dir <- asd_param_path(params$cache_dir, asd_db_snapshot())
dir.create(cache_dir, showWarnings = FALSE, recursive = TRUE)

rds_path      <- file.path(cache_dir, "df_cds.rds")
csv_path      <- file.path(cache_dir, "df_cds.csv")
raw_rds_path  <- file.path(cache_dir, "df_raw.rds.gz")   # nested GraphQL blob
manifest_path <- file.path(cache_dir, "manifest.json")
```

# Read-only safety

The Hasura admin secret has full RW power.  We never use it for writes —
this notebook only ever issues read queries.  Two belts:

1. `gql_read()` refuses anything that doesn't start with `query` or `{`,
   and rejects any query body containing `mutation`, `insert_`, `update_`,
   or `delete_` tokens.
2. Every call site uses `gql_read()` — never raw `POST`.

```{r gql_read_wrapper}
# Load HASURA_ADMIN_SECRET from the .env file if not already in env.
# We deliberately do NOT load HASURA_URL from .env — that file uses shell-style
# `${GRAPHQL_PORT}` substitution which R doesn't expand.  The URL always comes
# from the rmd's `params$hasura_url`.
.load_hasura_secret <- function(env_file) {
  if (Sys.getenv("HASURA_ADMIN_SECRET") != "") return(invisible(NULL))
  if (!file.exists(env_file)) return(invisible(NULL))
  lines <- readLines(env_file, warn = FALSE)
  for (l in lines) {
    l <- trimws(l)
    if (l == "" || startsWith(l, "#")) next
    m <- regmatches(l, regexec("^HASURA_ADMIN_SECRET=\"?([^\"]*)\"?$", l))[[1]]
    if (length(m) == 2) {
      Sys.setenv(HASURA_ADMIN_SECRET = m[2])
      return(invisible(NULL))
    }
  }
}

.load_hasura_secret(asd_param_path(params$env_file, asd_env_file()))

HASURA_URL    <- params$hasura_url                 # always from params
HASURA_SECRET <- Sys.getenv("HASURA_ADMIN_SECRET")
stopifnot(nchar(HASURA_SECRET) > 0)

# Read-only GraphQL caller.  REFUSES anything that isn't a pure read query.
gql_read <- function(query, variables = NULL, max_retries = 5) {
  q <- trimws(query)
  first <- regmatches(q, regexpr("^(query\\b|\\{)", q))
  if (length(first) == 0)
    stop("gql_read: query must start with 'query' or '{'.")
  q_lower <- tolower(q)
  for (tok in c("\nmutation ", " mutation ", "mutation{", "mutation \\{",
                "insert_", "update_", "delete_")) {
    if (grepl(tok, q_lower, fixed = !grepl("\\\\", tok))) {
      stop("gql_read: forbidden token '", trimws(tok), "' detected — refusing.")
    }
  }

  body <- list(query = query)
  if (!is.null(variables)) body$variables <- variables

  for (attempt in seq_len(max_retries)) {
    resp <- tryCatch(
      POST(
        url  = HASURA_URL,
        add_headers(`x-hasura-admin-secret` = HASURA_SECRET),
        content_type_json(),
        body = toJSON(body, auto_unbox = TRUE),
        timeout(300)
      ),
      # cat() not message() — chunks default to message=FALSE so message()
      # silently swallows retry errors, which makes debugging impossible.
      error = function(e) { cat("[gql_read] attempt ", attempt, " POST error: ",
                                conditionMessage(e), "\n", sep=""); NULL }
    )
    if (is.null(resp))                       { Sys.sleep(0.5 * 2^(attempt-1)); next }
    if (status_code(resp) %in% c(502,503,504)) {
      cat("[gql_read] attempt ", attempt, " HTTP ", status_code(resp), "\n", sep="")
      Sys.sleep(0.5 * 2^(attempt-1)); next
    }
    parsed <- fromJSON(content(resp, as = "text", encoding = "UTF-8"), flatten = TRUE)
    if (!is.null(parsed$errors))
      stop("GraphQL error: ", toJSON(parsed$errors, auto_unbox = TRUE))
    return(as.data.frame(parsed$data[[1]]))
  }
  stop("gql_read: failed after ", max_retries, " attempts")
}
```

# Cache check

If `regenerate` is FALSE and the cache exists, we just load it.  Otherwise
we pull from Hasura and overwrite the cache.

```{r cache_check}
have_cache <- file.exists(rds_path) && file.exists(manifest_path)

if (!params$regenerate && have_cache) {
  message("Reading cached snapshot from: ", rds_path)
  df_cds <- readRDS(rds_path)
  manifest <- fromJSON(manifest_path)
  message("  Pulled at:  ", manifest$pulled_at)
  message("  DB:         ", manifest$hasura_url)
  message("  Variants:   ", nrow(df_cds))
  message("  Genes:      ", length(unique(df_cds$gene_symbol)))
} else if (!params$regenerate && !have_cache) {
  stop("No cache found at ", rds_path,
       " — re-knit with params=list(regenerate=TRUE) to pull from DB.")
}
```

# Pull (only when `regenerate=TRUE`)

The query below is copied verbatim from `ASD_code_try_ver4_130326_Nave_dataset.rmd`
(the study team, March 2026).  Filters: SNV ∩ protein_coding
∩ gene has a SFARI Gene Score ∩ consequence ∈ {StopGained, Missense, Splice,
Frameshift, Synonymous, StartLost, StopRetained}.

```{r pull_raw, eval = params$regenerate}
message("Pulling from: ", HASURA_URL)

variants_query <- '
{
  variants(
    where: {
      class: { _eq: "SNV" }
      variants_features: {
        feature: { biotype: { name: { _eq: "protein_coding" } } }
      }
      gene: {
        genes_quantitative_scores: {
          quantitative_score: { name: { _eq: "SFARI Gene Score" } }
        }
      }
      variants_consequences: {
        consequence: { name: { _in: ["StopGained", "Missense", "Splice", "Frameshift", "Synonymous", "StartLost", "StopRetained"] } }
      }
    }
  ) {
    id
    ref
    alt
    class
    cds_position
    protein_position
    exon
    intron
    inheritance
    rs_id
    nmd_escaping_variant
    position_in_codon

    coordinate { chr start end strand }
    gene { ensg symbol strand }
    ref_codon { nt1 nt2 nt3 is_stop amino_acid { letter full_name } }
    alt_codon { nt1 nt2 nt3 is_stop amino_acid { letter full_name } }

    variants_consequences(
      where: { consequence: { name: { _in: ["StopGained", "Missense", "Splice", "Frameshift", "Synonymous", "StartLost", "StopRetained"] } } }
    ) {
      consequence { name }
    }

    gene {
      genes_quantitative_scores { quantitative_score { name value } }
    }
    gene {
      genes_qualitative_scores  { qualitative_score  { name value } }
    }

    variants_subjects { subject_id }
    variants_papers   { paper { paper_key } }

    variants_qualitative_scores  { qualitative_score  { name value } }
    variants_quantitative_scores { quantitative_score { name value } }
  }
}'

t0 <- Sys.time()
df_raw <- gql_read(variants_query)
message("Raw pull: ", nrow(df_raw), " rows in ",
        format(round(Sys.time() - t0, 1)))
```

# Flatten into df_cds

```{r flatten, eval = params$regenerate}
safe_extract <- function(lst, fn) {
  sapply(lst, function(x) tryCatch(fn(x), error = function(e) NA))
}

xq <- function(score_name, kind = c("var_qual","var_quant","gene_qual","gene_quant")) {
  kind <- match.arg(kind)
  col <- switch(kind,
    var_qual   = "variants_qualitative_scores",
    var_quant  = "variants_quantitative_scores",
    gene_qual  = "gene.genes_qualitative_scores",
    gene_quant = "gene.genes_quantitative_scores")
  is_qual <- endsWith(kind, "qual")
  field   <- if (is_qual) "qualitative_score.name" else "quantitative_score.name"
  valfld  <- if (is_qual) "qualitative_score.value" else "quantitative_score.value"

  safe_extract(df_raw[[col]], function(x) {
    if (is.null(x) || nrow(x) == 0) return(if (is_qual) NA_character_ else NA_real_)
    m <- x[x[[field]] == score_name, , drop = FALSE]
    if (nrow(m) == 0) return(if (is_qual) NA_character_ else NA_real_)
    if (is_qual) paste(m[[valfld]], collapse = "&") else as.numeric(m[[valfld]][1])
  })
}

df_cds <- data.frame(
  id                   = df_raw$id,
  REF.VEP              = df_raw$ref,
  ALT.VEP              = df_raw$alt,
  class                = df_raw$class,
  cds_position         = df_raw$cds_position,
  protein_position     = df_raw$protein_position,
  exon                 = df_raw$exon,
  intron               = df_raw$intron,
  inheritance          = df_raw$inheritance,
  rs_id                = df_raw$rs_id,
  nmd_escaping_variant = df_raw$nmd_escaping_variant,
  position_in_codon    = df_raw$position_in_codon,
  stringsAsFactors = FALSE
)

df_cds$chr    <- df_raw[["coordinate.chr"]]
df_cds$start  <- df_raw[["coordinate.start"]]
df_cds$end    <- df_raw[["coordinate.end"]]
df_cds$STRAND <- df_raw[["coordinate.strand"]]

df_cds$gene_ensg   <- df_raw[["gene.ensg"]]
df_cds$gene_symbol <- df_raw[["gene.symbol"]]
df_cds$gene_strand <- df_raw[["gene.strand"]]

df_cds$ref_codon_nt1  <- df_raw[["ref_codon.nt1"]]
df_cds$ref_codon_nt2  <- df_raw[["ref_codon.nt2"]]
df_cds$ref_codon_nt3  <- df_raw[["ref_codon.nt3"]]
df_cds$ref_codon_stop <- df_raw[["ref_codon.is_stop"]]
df_cds$ref_aa         <- df_raw[["ref_codon.amino_acid.letter"]]
df_cds$ref_aa_name    <- df_raw[["ref_codon.amino_acid.full_name"]]

df_cds$alt_codon_nt1  <- df_raw[["alt_codon.nt1"]]
df_cds$alt_codon_nt2  <- df_raw[["alt_codon.nt2"]]
df_cds$alt_codon_nt3  <- df_raw[["alt_codon.nt3"]]
df_cds$alt_codon_stop <- df_raw[["alt_codon.is_stop"]]
df_cds$alt_aa         <- df_raw[["alt_codon.amino_acid.letter"]]
df_cds$alt_aa_name    <- df_raw[["alt_codon.amino_acid.full_name"]]

df_cds$Consequence <- safe_extract(df_raw[["variants_consequences"]], function(x) {
  if (is.null(x) || nrow(x) == 0) return(NA_character_)
  # Sorted, not in arrival order: the database returns a variant's consequences in
  # no guaranteed order, so an unsorted join makes this column -- and therefore the
  # whole snapshot -- differ between identical pulls. 243 of the 9,962 rows carry
  # more than one consequence and were flipping between runs. Every consumer tests
  # membership (str_detect / grepl), never the string itself, so sorting changes no
  # result; it only makes the file reproducible.
  paste(sort(x$consequence.name), collapse = ",")
})

df_cds$New_Consequence <- dplyr::case_when(
  grepl("StopGained",   df_cds$Consequence) ~ "Nonsense",
  grepl("Splice",       df_cds$Consequence) ~ "Splice",
  grepl("Missense",     df_cds$Consequence) ~ "Missense",
  grepl("Frameshift",   df_cds$Consequence) ~ "Frameshift",
  grepl("Synonymous",   df_cds$Consequence) ~ "Synonymous",
  grepl("StartLost",    df_cds$Consequence) ~ "Start_lost",
  grepl("StopRetained", df_cds$Consequence) ~ "Stop_retained",
  TRUE                                      ~ "Other"
)

df_cds$patient_count <- safe_extract(df_raw[["variants_subjects"]], function(x) {
  if (is.null(x) || nrow(x) == 0) return(0L)
  length(unique(x$subject_id))
})

df_cds$paper_keys <- safe_extract(df_raw[["variants_papers"]], function(x) {
  if (is.null(x) || nrow(x) == 0) return(NA_character_)
  paste(x$paper.paper_key, collapse = ";")
})

df_cds$CLIN_SIG      <- xq("ClinVar",       "var_qual")
df_cds$SIFT_pred     <- xq("SIFT",          "var_qual")
df_cds$am_class      <- xq("AlphaMissense", "var_qual")
df_cds$condel_pred   <- xq("Condel",        "var_qual")
df_cds$polyphen_pred <- xq("PolyPhen",      "var_qual")

df_cds$SIFT_score       <- xq("SIFT",          "var_quant")
df_cds$am_pathogenicity <- xq("AlphaMissense", "var_quant")
df_cds$polyphen_score   <- xq("PolyPhen",      "var_quant")
df_cds$REVEL            <- xq("REVEL",         "var_quant")
df_cds$BLOSUM62         <- xq("BLOSUM62",      "var_quant")
df_cds$condel_score     <- xq("Condel",        "var_quant")
df_cds$pLI              <- xq("pLI",           "var_quant")
df_cds$PhastCons100     <- xq("PhastCons100",  "var_quant")
df_cds$LoFtool          <- xq("LoFtool",       "var_quant")
df_cds$GENE_PHENO       <- xq("GENE_PHENO",    "var_quant")
df_cds$CADD_phred       <- xq("CADD_phred",    "var_quant")

df_cds$sfari_syndromic  <- xq("SFARI Syndromic", "gene_qual")
df_cds$sfari_gene_score <- xq("SFARI Gene Score","gene_quant")
df_cds$sfari_reports    <- xq("SFARI #Reports",  "gene_quant")
```

# Strand correction

VEP reports REF/ALT on the chromosome (+) strand.  A G>A on a minus-strand
gene is reported as C>T.  This block creates strand-corrected `REF` / `ALT`
columns, keeping the originals as `REF.VEP` / `ALT.VEP`.

```{r strand_correct, eval = params$regenerate}
complement_base <- function(b) chartr("ACGT", "TGCA", b)

df_cds <- df_cds %>%
  mutate(
    REF = ifelse(gene_strand == "-", complement_base(REF.VEP), REF.VEP),
    ALT = ifelse(gene_strand == "-", complement_base(ALT.VEP), ALT.VEP)
  )

# Sanity print
ga_before <- sum(df_cds$REF.VEP == "G" & df_cds$ALT.VEP == "A", na.rm = TRUE)
ga_after  <- sum(df_cds$REF     == "G" & df_cds$ALT     == "A", na.rm = TRUE)
ct_before <- sum(df_cds$REF.VEP == "C" & df_cds$ALT.VEP == "T", na.rm = TRUE)
ct_after  <- sum(df_cds$REF     == "C" & df_cds$ALT     == "T", na.rm = TRUE)
message("G>A before/after strand fix: ", ga_before, " → ", ga_after)
message("C>T before/after strand fix: ", ct_before, " → ", ct_after)
```

# Save snapshot + manifest

```{r save_cache, eval = params$regenerate}
saveRDS(df_cds, rds_path)                              # default = gzip
write.csv(df_cds, csv_path, row.names = FALSE)
saveRDS(df_raw, raw_rds_path, compress = "gzip")       # full nested GraphQL blob

# Content first, then the run context. The split is deliberate: everything above
# `pulled_at` is a fact about the snapshot and is identical between two honest
# pulls; everything below records the circumstances of one run and is not.
# `scripts/verify_counts.py` checksums the content half only, so a re-run does not
# look like a regression while the facts still cannot drift unnoticed.
#
# Paths are recorded relative to the output root -- they used to be absolute,
# which wrote one machine's directory layout into a file a reader may well see.
manifest <- list(
  query_sha1     = digest::digest(variants_query, algo = "sha1"),
  n_variants     = nrow(df_cds),
  n_genes        = length(unique(df_cds$gene_symbol)),
  n_columns      = ncol(df_cds),
  consequence_counts = as.list(table(df_cds$New_Consequence, useNA = "ifany")),
  raw_rds_bytes  = file.info(raw_rds_path)$size,
  rds_file       = basename(rds_path),
  csv_file       = basename(csv_path),
  raw_rds_file   = basename(raw_rds_path),
  pulled_at      = format(Sys.time(), tz = "UTC", usetz = TRUE),
  hasura_url     = HASURA_URL
)
writeLines(toJSON(manifest, pretty = TRUE, auto_unbox = TRUE), manifest_path)

message("\nCache written:")
message("  ", rds_path)
message("  ", csv_path)
message("  ", raw_rds_path,
        " (", format(file.info(raw_rds_path)$size, big.mark = ","), " bytes)")
message("  ", manifest_path)
```

# Sanity report

```{r sanity}
cat("Variants:   ", nrow(df_cds), "\n")
cat("Genes:      ", length(unique(df_cds$gene_symbol)), "\n")
cat("\nConsequence breakdown:\n")
print(table(df_cds$New_Consequence, useNA = "ifany"))
cat("\nMismatch breakdown (top 10, strand-corrected):\n")
print(sort(table(paste0(df_cds$REF, ">", df_cds$ALT)), decreasing = TRUE)[1:10])
cat("\nSFARI gene score:\n")
print(table(df_cds$sfari_gene_score, useNA = "ifany"))
cat("\nClinVar coverage: ", sum(!is.na(df_cds$CLIN_SIG)), " / ", nrow(df_cds), "\n")
```

# Session info

```{r}
sessionInfo()
```
