---
title: "Figure 1A — Variant Filtering Funnel"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
params:
  regenerate:        false                                                      # FALSE: reload cached counts
  panel_id:          "Figure_1A_funnel"
  hasura_url:       "http://localhost:8789/v1/graphql"
  env_file:         ""
  intermediate_dir: ""
  out_main:         ""
  out_supp_tables:  ""
  width:             6
  height:            8
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

root <- Sys.getenv("ASD_PAPER_ROOT")
if (!nzchar(root)) stop("ASD_PAPER_ROOT is unset -- see figures/README.md")
source(file.path(root, "figures", "_paths.R"))

suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(scales)
  library(httr)
  library(jsonlite)
})

source(asd_code("ver1", "_helpers.R"))  # theme_publication, fig_colors, save_panel

panel_id        <- params$panel_id
intermediate_dir<- asd_param_path(params$intermediate_dir, asd_intermediate_dir())
out_main        <- asd_param_path(params$out_main,         asd_panels_dir())
out_supp_tables <- asd_param_path(params$out_supp_tables,  asd_supp_tables_dir("panels"))
dir.create(intermediate_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(out_main,         showWarnings = FALSE, recursive = TRUE)
dir.create(out_supp_tables,  showWarnings = FALSE, recursive = TRUE)

counts_rds       <- file.path(intermediate_dir, paste0(panel_id, "_counts.rds"))
ggplot_rds       <- file.path(intermediate_dir, paste0(panel_id, "_ggplot.rds"))
source_data_csv  <- file.path(intermediate_dir, paste0("Source_data_", panel_id, ".csv"))
```

# Plan

Sequential filter funnel — 5 levels of narrowing from VariCarta total to
the CDS+Splice subset that drives every other Figure 1 panel.  Linear
top-to-bottom layout ("simple list", per the collaborators' spec); each box shows
the category, count, and percentage retained from the previous step.
Drop annotations on the right of each arrow show how many variants were
filtered out at each stage.

Five filter steps (matches the exploratory notebook's main path, dropping
his branch siblings):

| step | filter |
|---:|---|
| 1 | All VariCarta variants |
| 2 | SNV only |
| 3 | + protein-coding biotype |
| 4 | + gene has SFARI Gene Score |
| 5 | + consequence ∈ {StopGained, Missense, Splice, Frameshift, Synonymous, StartLost, StopRetained} |

Step 5's row count (9,632) is exactly the variant count cached by
`00_prep_data.rmd`.

# Read-only GraphQL wrapper

```{r gql_wrapper}
.load_hasura_secret <- function(env_file) {
  if (Sys.getenv("HASURA_ADMIN_SECRET") != "") return(invisible(NULL))
  if (!file.exists(env_file)) return(invisible(NULL))
  for (l in readLines(env_file, warn = FALSE)) {
    l <- trimws(l)
    m <- regmatches(l, regexec("^HASURA_ADMIN_SECRET=\"?([^\"]*)\"?$", l))[[1]]
    if (length(m) == 2) { Sys.setenv(HASURA_ADMIN_SECRET = m[2]); break }
  }
}
.load_hasura_secret(asd_param_path(params$env_file, asd_env_file()))

HASURA_URL    <- params$hasura_url
HASURA_SECRET <- Sys.getenv("HASURA_ADMIN_SECRET")
stopifnot(nchar(HASURA_SECRET) > 0)

gql_read <- function(query, max_retries = 5) {
  q <- trimws(query)
  if (!grepl("^(query\\b|\\{)", q))
    stop("gql_read: query must start with 'query' or '{'.")
  for (tok in c("mutation", "insert_", "update_", "delete_"))
    if (grepl(tok, tolower(q), fixed = TRUE))
      stop("gql_read: forbidden token '", tok, "' detected.")

  for (attempt in seq_len(max_retries)) {
    resp <- tryCatch(
      POST(HASURA_URL,
           add_headers(`x-hasura-admin-secret` = HASURA_SECRET),
           content_type_json(),
           body = toJSON(list(query = query), auto_unbox = TRUE),
           timeout(120)),
      error = function(e) { cat("[gql_read] err: ", conditionMessage(e), "\n", sep=""); NULL })
    if (is.null(resp)) { Sys.sleep(0.5 * 2^(attempt-1)); next }
    parsed <- fromJSON(content(resp, as = "text", encoding = "UTF-8"), flatten = TRUE)
    if (!is.null(parsed$errors))
      stop("GraphQL error: ", toJSON(parsed$errors, auto_unbox = TRUE))
    return(parsed$data)
  }
  stop("gql_read: failed after ", max_retries, " attempts")
}
```

# Pull (or reload) the 5 counts

```{r pull_counts}
have_cache <- file.exists(counts_rds)

if (!params$regenerate && have_cache) {
  steps <- readRDS(counts_rds)
  cat("reusing cached counts:", counts_rds, "\n")
} else {
  query <- '
  {
    s1_total: variants_aggregate { aggregate { count } }

    s2_snv: variants_aggregate(where: { class: { _eq: "SNV" } }) {
      aggregate { count }
    }

    s3_pc: variants_aggregate(where: {
      class: { _eq: "SNV" }
      variants_features: { feature: { biotype: { name: { _eq: "protein_coding" } } } }
    }) { aggregate { count } }

    s4_sfari: variants_aggregate(where: {
      class: { _eq: "SNV" }
      variants_features: { feature: { biotype: { name: { _eq: "protein_coding" } } } }
      gene: { genes_quantitative_scores: { quantitative_score: { name: { _eq: "SFARI Gene Score" } } } }
    }) { aggregate { count } }

    s5_cds: variants_aggregate(where: {
      class: { _eq: "SNV" }
      variants_features: { feature: { biotype: { name: { _eq: "protein_coding" } } } }
      gene: { genes_quantitative_scores: { quantitative_score: { name: { _eq: "SFARI Gene Score" } } } }
      variants_consequences: { consequence: { name: { _in: ["StopGained", "Missense", "Splice", "Frameshift", "Synonymous", "StartLost", "StopRetained"] } } }
    }) { aggregate { count } }
  }'

  d <- gql_read(query)

  steps <- data.frame(
    idx   = 1:5,
    title = c("Total variants (VariCarta)",
              "Single nucleotide variants\n(SNVs)",      # two lines
              "Protein-coding regions",
              "In SFARI genes",
              "CDS + splice sites"),
    count = c(d$s1_total$aggregate$count,
              d$s2_snv$aggregate$count,
              d$s3_pc$aggregate$count,
              d$s4_sfari$aggregate$count,
              d$s5_cds$aggregate$count),
    stringsAsFactors = FALSE
  )
  steps$pct_of_total <- 100 * steps$count / steps$count[1]
  steps$pct_of_prev  <- c(NA, 100 * steps$count[-1] / steps$count[-nrow(steps)])
  steps$dropped      <- c(NA, steps$count[-nrow(steps)] - steps$count[-1])
  steps$y            <- -steps$idx

  saveRDS(steps, counts_rds)
  write.csv(steps, source_data_csv, row.names = FALSE)
}

print(steps)
```

# Build the funnel

```{r build_panel, fig.width=params$width, fig.height=params$height}
# Box-less funnel: text only with arrows between steps.
# `half_w` is retained purely to define the x-axis extent.
# ggplot geom_text `size` is in mm; size=5 ≈ 14 pt visually.
half_w  <- 2.0
y_scale <- 1.00   # vertical spacing between text blocks

boxes <- steps %>%
  mutate(
    y           = y * y_scale,
    label_main  = title,
    label_count = format(count, big.mark = ",")
  )

# Arrows between consecutive text blocks
arrows <- data.frame(
  x    = 0,
  xend = 0,
  y    = head(boxes$y, -1) - 0.22,   # below count text of upper step
  yend = tail(boxes$y, -1) + 0.22    # above title text of lower step
)

p <- ggplot() +
  # Title text (top line)
  geom_text(
    data = boxes,
    aes(x = 0, y = y + 0.12, label = label_main),
    family = "Nimbus Sans", fontface = "bold", size = 5,
    lineheight = 0.85
  ) +
  # Count text (bottom line)
  geom_text(
    data = boxes,
    aes(x = 0, y = y - 0.12, label = label_count),
    family = "Nimbus Sans", fontface = "plain", size = 5,
    colour = "grey25"
  ) +
  # Arrows
  geom_segment(
    data = arrows,
    aes(x = x, xend = xend, y = y, yend = yend),
    arrow = arrow(length = unit(0.20, "cm"), type = "closed"),
    linewidth = 0.6
  ) +
  scale_x_continuous(limits = c(-half_w - 0.4, half_w + 2.4),
                     expand = expansion(0)) +
  scale_y_continuous(limits = c(min(boxes$y) - 0.4, max(boxes$y) + 0.4),
                     expand = expansion(0)) +
  coord_cartesian(clip = "off") +
  labs(tag = "A") +
  theme_void(base_family = "Nimbus Sans") +
  theme(
    plot.tag           = element_text(face = "bold", size = 16, family = "Nimbus Sans"),
    plot.tag.position  = c(0, 1),
    plot.background    = element_blank(),
    panel.background   = element_blank(),
    plot.margin        = margin(20, 10, 10, 10)
  )

print(p)
```

# Save (PDF + ggplot RDS)

```{r save}
save_panel(p, panel_id,
           pdf_dir = out_main,
           rds_dir = intermediate_dir,
           width  = params$width,
           height = params$height)

write.csv(steps,
          file.path(out_supp_tables, paste0("Supp_Table_F1A_", panel_id, ".csv")),
          row.names = FALSE)
```

# Session info

```{r}
sessionInfo()
```
