---
title: "Figure 1B — Mismatch Type Distribution"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
params:
  regenerate:        false                                                      # FALSE: reload cached ggplot, just re-save PDF
  panel_id:          "Figure_1B_mismatch"
  cache_db:         ""
  intermediate_dir: ""
  out_main:         ""
  out_supp_tables:  ""
  width:             8
  height:            5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

root <- Sys.getenv("ASD_PAPER_ROOT")
if (!nzchar(root)) stop("ASD_PAPER_ROOT is unset -- see figures/README.md")
source(file.path(root, "figures", "_paths.R"))

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(scales)
  library(stringr)
})

# ggrepel resolves label collisions with random jitter, so an unseeded run places
# domain labels differently every time. Nothing in the data moves -- only where a
# label lands and how its leader line is drawn -- but it is the reason these
# figures were not byte-reproducible. Seeded here so they are.
set.seed(42)

source(asd_code("ver1", "_helpers.R"))  # theme_publication, fig_colors, mismatch_fill, consequence_fill, save_panel

# Resolve paths (relative to this rmd)
here_rmd <- function(p) normalizePath(p, mustWork = FALSE)
panel_id        <- params$panel_id
db_cache_path   <- asd_param_path(params$cache_db, asd_db_snapshot("df_cds.rds"))
intermediate_dir<- asd_param_path(params$intermediate_dir, asd_intermediate_dir())
out_main        <- asd_param_path(params$out_main, asd_panels_dir())
out_supp_tables <- asd_param_path(params$out_supp_tables, asd_supp_tables_dir("panels"))
dir.create(intermediate_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(out_main,         showWarnings = FALSE, recursive = TRUE)
dir.create(out_supp_tables,  showWarnings = FALSE, recursive = TRUE)

plot_data_rds   <- file.path(intermediate_dir, paste0(panel_id, "_plotdata.rds"))
ggplot_rds      <- file.path(intermediate_dir, paste0(panel_id, "_ggplot.rds"))
source_data_csv <- file.path(intermediate_dir, paste0("Source_data_", panel_id, ".csv"))
totals_csv      <- file.path(intermediate_dir, paste0("Totals_",      panel_id, ".csv"))
```

# Plan

Stacked bar of variant counts per mismatch type (12 types).  Within each
bar, segments coloured by consequence — Missense / Nonsense / Splice /
Synonymous / Start Lost.  Bars sorted by total count, descending.  Total
count printed above each bar.

The ADAR-treatable mismatch is **G>A** on the gene-sense strand (it
becomes A>G on the mRNA, then ADAR edits A→I).  Strand correction was
already applied in `00_prep_data.rmd`, so `REF`/`ALT` here are
gene-sense; G>A is what we expect to see prominent.

# Read cached snapshot

```{r read_cache}
stopifnot(file.exists(db_cache_path))
df_cds <- readRDS(db_cache_path)
cat("loaded:", db_cache_path, "—", nrow(df_cds), "variants\n")
```

# Compute plot data (or reload)

If `regenerate=FALSE` and the plot-data cache exists, just reload it —
this lets us tweak colours / theme without re-running the count.

```{r build_plot_data}
have_plotdata <- file.exists(plot_data_rds)

if (!params$regenerate && have_plotdata) {
  cat("reusing cached plot data:", plot_data_rds, "\n")
  bundle <- readRDS(plot_data_rds)
  plot_data <- bundle$plot_data
  totals    <- bundle$totals
} else {
  # SNVs only (single base on each side) — REF/ALT are already strand-corrected
  plot_data <- df_cds %>%
    filter(nchar(REF) == 1L, nchar(ALT) == 1L) %>%
    mutate(
      Mutation_Type = paste0(REF, ">", ALT),
      first_cons    = sapply(str_split(New_Consequence, ","), `[`, 1),
      Simple_consequence = case_when(
        first_cons == "Nonsense"   ~ "Nonsense",
        first_cons == "Splice"     ~ "Splice",
        first_cons == "Missense"   ~ "Missense",
        first_cons == "Synonymous" ~ "Synonymous",
        first_cons == "Start_lost" ~ "Start Lost",
        TRUE                       ~ NA_character_
      )
    ) %>%
    filter(!is.na(Simple_consequence)) %>%
    count(Mutation_Type, Simple_consequence, name = "n")

  totals <- plot_data %>%
    group_by(Mutation_Type) %>%
    summarise(Total_N = sum(n), .groups = "drop") %>%
    arrange(desc(Total_N))

  # Lock factor orders so the plot is reproducible from the rds
  mt_order <- totals$Mutation_Type
  cons_order <- c("Missense", "Nonsense", "Splice", "Synonymous", "Start Lost")
  plot_data$Mutation_Type      <- factor(plot_data$Mutation_Type,      levels = mt_order)
  plot_data$Simple_consequence <- factor(plot_data$Simple_consequence, levels = cons_order)
  totals$Mutation_Type         <- factor(totals$Mutation_Type,         levels = mt_order)

  saveRDS(list(plot_data = plot_data, totals = totals), plot_data_rds)
  write.csv(plot_data, source_data_csv, row.names = FALSE)
  write.csv(totals,    totals_csv,      row.names = FALSE)
  cat("wrote:", plot_data_rds, "\n")
  cat("wrote:", source_data_csv, "\n")
}

print(totals)
```

# Build the panel

If a cached ggplot RDS exists and `regenerate=FALSE`, we reload it
(useful when downstream composers re-run the figure) — *unless* you've
edited the helpers, in which case rebuild from `plot_data`.

```{r build_panel, fig.width=params$width, fig.height=params$height}
have_ggplot <- file.exists(ggplot_rds)

# Always rebuild from plot_data when we just recomputed it; otherwise
# rebuild anyway so palette edits in _helpers.R land.  Reloading the RDS
# is only useful for the composer, not for in-panel iteration.
p <- ggplot(plot_data,
            aes(x = Mutation_Type, y = n, fill = Simple_consequence)) +
  geom_col(width = 0.75, colour = "black", linewidth = 0.3) +
  geom_text(
    data = totals,
    aes(x = Mutation_Type, y = Total_N, label = format(Total_N, big.mark = ",")),
    inherit.aes = FALSE,
    vjust = -0.4, fontface = "bold", size = 3.5,
    family = "Nimbus Sans"
  ) +
  scale_fill_manual(values = consequence_fill, name = NULL,
                    breaks = c("Missense", "Nonsense", "Splice",
                               "Synonymous", "Start Lost")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)),
                     labels = comma) +
  labs(x = "mismatch type", y = "number of variants",
       tag = "B") +
  theme_publication(base_size = 12, x_angle = 0, y_bold = TRUE) +
  theme(
    legend.position    = "top",
    legend.justification= "center",
    panel.grid.major.x = element_blank(),
    plot.tag           = element_text(face = "bold", size = 16,
                                      family = "Nimbus Sans"),
    plot.tag.position  = c(0, 1)
  )

print(p)
```

# Save (PDF + ggplot RDS)

```{r save}
save_panel(p, panel_id,
           pdf_dir = out_main,
           rds_dir = intermediate_dir,
           width  = params$width,
           height = params$height)
```

# G>A-only version: single bar with in-segment count labels

Requested by the collaborators: a focused version of Panel B that drops the other 11
mismatches and shows only the G>A column.  Counts are written **inside
each consequence rectangle** (the bar's segments).  For segments too
small to hold a label (e.g. Start Lost = 1), `ggrepel` puts the label
outside with a leader line.

```{r panel_b_ga_only, fig.width=8, fig.height=2.4}
suppressPackageStartupMessages(library(ggrepel))

ga_only <- plot_data %>%
  filter(Mutation_Type == "G>A") %>%
  # Custom order so leader-label segments (Start Lost, Nonsense) sit at the
  # LEFT and in-segment-label segments (Splice, Synonymous) sit at the RIGHT,
  # with giant Missense anchoring the middle.  Separates the two label
  # styles into different zones so they don't crowd each other.
  arrange(match(as.character(Simple_consequence),
                c("Start Lost", "Nonsense", "Missense", "Splice", "Synonymous")))
ga_only$xmax <- cumsum(ga_only$n)
ga_only$xmin <- ga_only$xmax - ga_only$n
ga_only$xmid <- (ga_only$xmin + ga_only$xmax) / 2
ga_only$label <- format(ga_only$n, big.mark = ",")

# Inside label: "Consequence\nN" (legend dropped)
ga_only$inside_label <- paste0(ga_only$Simple_consequence, "\n", ga_only$label)

ga_total <- sum(ga_only$n)
# All consequence labels go ABOVE the bar via ggrepel — no in-segment text.
ga_only$inside <- FALSE

# Build a small dataframe for the above-the-bar leader labels.
get_n    <- function(nm) ga_only$n   [as.character(ga_only$Simple_consequence) == nm]
get_xmid <- function(nm) ga_only$xmid[as.character(ga_only$Simple_consequence) == nm]
get_xmin <- function(nm) ga_only$xmin[as.character(ga_only$Simple_consequence) == nm]
get_xmax <- function(nm) ga_only$xmax[as.character(ga_only$Simple_consequence) == nm]

above_labels <- data.frame(
  x = c(
    (get_xmin("Start Lost") + get_xmax("Nonsense")) / 2,
    get_xmid("Missense"),
    get_xmid("Splice"),
    get_xmid("Synonymous")
  ),
  label = c(
    sprintf("Nonsense: %s\nStart Lost: %s",
            format(get_n("Nonsense"),   big.mark = ","),
            format(get_n("Start Lost"), big.mark = ",")),
    sprintf("Missense: %s",   format(get_n("Missense"),   big.mark = ",")),
    sprintf("Splice: %s",     format(get_n("Splice"),     big.mark = ",")),
    sprintf("Synonymous: %s", format(get_n("Synonymous"), big.mark = ","))
  ),
  stringsAsFactors = FALSE
)

# Grayscale by severity: light (Synonymous) -> near black (Start Lost).
# Strictly monochrome for B&W publication; ordering follows the severity
# ranking used in the earlier colour palette.
ga_consequence_fill <- c(
  "Synonymous" = "#EAEAEA",   # very light gray  (mild)
  "Missense"   = "#BFBFBF",   # light-medium
  "Splice"     = "#888888",   # medium
  "Nonsense"   = "#4D4D4D",   # dark gray
  "Start Lost" = "#1A1A1A"    # near black (most severe)
)
# All labels live ABOVE the bar (no in-segment text) so black on white is fine.
ga_only$text_colour <- "black"

# Immediately before the repel call, not at the top of the file: any code
# between the two advances the RNG, so a distant seed does not pin the
# label layout. This is why an earlier seeding attempt fixed Figure S1 but
# left Figures 1 and 3 varying run to run.
set.seed(42)
p_ga <- ggplot(ga_only) +
  geom_rect(
    aes(xmin = xmin, xmax = xmax, ymin = -0.4, ymax = 0.4,
        fill = Simple_consequence),
    colour = "black", linewidth = 0.4
  ) +
  # In-segment labels
  geom_text(
    data = subset(ga_only, inside),
    aes(x = xmid, y = 0, label = inside_label, colour = I(text_colour)),
    fontface = "bold", size = 4, family = "Nimbus Sans",
    lineheight = 0.9
  ) +
  # Outside labels with leader lines above the bar:
  #   - Nonsense + Start Lost combined into one two-line label
  #   - Splice and Synonymous each as their own label (avoids in-segment overlap)
  geom_text_repel(seed = 42, 
    data = above_labels,
    aes(x = x, y = 0.4, label = label),
    nudge_y = 0.75, hjust = 0.5, direction = "x",
    size = 3.6, family = "Nimbus Sans", fontface = "bold",
    lineheight = 0.9,
    segment.colour = "grey40", segment.size = 0.4,
    min.segment.length = 0, box.padding = 0.35
  ) +
  scale_fill_manual(values = ga_consequence_fill, guide = "none") +
  # Extra right padding so the cluster of small segments + outside labels
  # at the bar's right end has breathing room.
  scale_x_continuous(expand = expansion(mult = c(0.02, 0.30)),
                     labels = comma) +
  scale_y_continuous(limits = c(-1.4, 1.7), expand = expansion(0)) +
  labs(x = NULL, y = NULL, tag = "C") +
  theme_publication(base_size = 12, x_angle = 0, y_bold = TRUE) +
  theme(
    legend.position    = "none",
    axis.text.x        = element_blank(),
    axis.text.y        = element_blank(),
    axis.ticks.x       = element_blank(),
    axis.ticks.y       = element_blank(),
    axis.line.x        = element_blank(),
    axis.line.y        = element_blank(),
    axis.title.x       = element_blank(),
    axis.title.y       = element_blank(),
    panel.grid.major   = element_blank(),
    plot.tag           = element_text(face = "bold", size = 16,
                                      family = "Nimbus Sans"),
    plot.tag.position  = c(0, 1)
  )
print(p_ga)

save_panel(p_ga, paste0(panel_id, "_GA_only"),
           pdf_dir = out_main,
           rds_dir = intermediate_dir,
           width  = 8,
           height = 2.4)
```

# Companion supp table (wide)

```{r supp_table}
wide <- plot_data %>%
  pivot_wider(names_from  = Simple_consequence,
              values_from = n,
              values_fill = 0) %>%
  mutate(Total = rowSums(across(where(is.numeric)))) %>%
  arrange(desc(Total))

write.csv(wide,
          file.path(out_supp_tables, paste0("Supp_Table_F1B_", panel_id, "_wide.csv")),
          row.names = FALSE)
print(wide)
```

# Session info

```{r}
sessionInfo()
```
