---
title: "Figure 1C — ClinVar Donuts (SFARI × ClinVar grid)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
params:
  regenerate:        false                                                      # FALSE: reload cached plot data
  panel_id:          "Figure_1C_clinvar_donut"
  cache_db:         ""
  intermediate_dir: ""
  out_main:         ""
  out_supp_tables:  ""
  width:             7
  height:            6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

root <- Sys.getenv("ASD_PAPER_ROOT")
if (!nzchar(root)) stop("ASD_PAPER_ROOT is unset -- see figures/README.md")
source(file.path(root, "figures", "_paths.R"))

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(scales)
  library(scatterpie)
  library(ggforce)     # geom_circle — donut holes that scale with the pie
})

source(asd_code("ver1", "_helpers.R"))  # theme_publication, save_panel

panel_id        <- params$panel_id
db_cache_path   <- asd_param_path(params$cache_db,         asd_db_snapshot("df_cds.rds"))
intermediate_dir<- asd_param_path(params$intermediate_dir, asd_intermediate_dir())
out_main        <- asd_param_path(params$out_main,         asd_panels_dir())
out_supp_tables <- asd_param_path(params$out_supp_tables,  asd_supp_tables_dir("panels"))
dir.create(intermediate_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(out_main,         showWarnings = FALSE, recursive = TRUE)
dir.create(out_supp_tables,  showWarnings = FALSE, recursive = TRUE)

plot_data_rds   <- file.path(intermediate_dir, paste0(panel_id, "_plotdata.rds"))
ggplot_rds      <- file.path(intermediate_dir, paste0(panel_id, "_ggplot.rds"))
source_data_csv <- file.path(intermediate_dir, paste0("Source_data_", panel_id, ".csv"))
```

# Plan

3×3 grid of ClinVar donut pies.

- **Y-axis**: SFARI gene score (1 = high-confidence ASD gene, 2 = strong, 3 = suggestive).
- **X-axis**: ClinVar classification — Pathogenic / VUS / Benign.  (Per
  the 11 Feb meeting: only three groups, no syndromic split.)
- Each donut's **area** is proportional to the variant count in that bucket
  (radius = √N normalized).
- Each donut is split into **G>A** (the ADAR-treatable mismatch, in red2)
  vs **other** (white).  The redder the donut → the more ADAR-treatable
  variants in that SFARI × ClinVar cell.
- Center label: total variant count in that bucket.

ClinVar grouping rules (matches v4 RMD):

| group | rule |
|---|---|
| Pathogenic | contains `"pathogenic"` AND NOT `"conflict"` AND NOT `"benign"` |
| VUS        | contains `"uncertain"` |
| Benign     | contains `"benign"` AND NOT `"pathogenic"` |

Variants whose `CLIN_SIG` doesn't match any rule are dropped (they're
"NA / Drug response / Other_assertion" rows).

# Read cached snapshot

```{r read_cache}
stopifnot(file.exists(db_cache_path))
df_cds <- readRDS(db_cache_path)
cat("loaded:", db_cache_path, "—", nrow(df_cds), "variants\n")
cat("with SFARI 1-3:", sum(df_cds$sfari_gene_score %in% c(1,2,3), na.rm = TRUE), "\n")
cat("with ClinVar string:", sum(!is.na(df_cds$CLIN_SIG)), "\n")
```

# Compute plot data (or reload)

```{r build_plot_data}
have_plotdata <- file.exists(plot_data_rds)

if (!params$regenerate && have_plotdata) {
  cat("reusing cached plot data:", plot_data_rds, "\n")
  bundle <- readRDS(plot_data_rds)
  pie_plot_data   <- bundle$pie_plot_data
  pie_base_data   <- bundle$pie_base_data
  category_levels <- bundle$category_levels
} else {
  pie_base_data <- df_cds %>%
    filter(sfari_gene_score %in% c(1, 2, 3)) %>%
    mutate(
      Category = case_when(
        grepl("pathogenic", CLIN_SIG, ignore.case = TRUE) &
          !grepl("conflict|benign", CLIN_SIG, ignore.case = TRUE) ~ "Pathogenic",
        grepl("uncertain",  CLIN_SIG, ignore.case = TRUE)         ~ "VUS",
        grepl("benign",     CLIN_SIG, ignore.case = TRUE) &
          !grepl("pathogenic", CLIN_SIG, ignore.case = TRUE)      ~ "Benign",
        TRUE                                                      ~ NA_character_
      ),
      Mutation_Type = ifelse(REF == "G" & ALT == "A", "G>A", "other")
    ) %>%
    filter(!is.na(Category))

  category_levels <- c("Pathogenic", "VUS", "Benign")

  pie_plot_data <- pie_base_data %>%
    group_by(sfari_gene_score, Category, Mutation_Type) %>%
    summarise(Count = n(), .groups = "drop") %>%
    pivot_wider(names_from = Mutation_Type, values_from = Count, values_fill = 0) %>%
    {                                              # ensure both pie cols always exist
      if (!"G>A"   %in% names(.)) mutate(., `G>A` = 0) else .
    } %>%
    { if (!"other" %in% names(.)) mutate(., other = 0) else . } %>%
    mutate(
      Total  = `G>A` + other,
      radius = sqrt(Total) / max(sqrt(Total)) * 0.45,
      x_pos  = as.numeric(factor(Category, levels = category_levels)),
      y_pos  = sfari_gene_score
    )

  saveRDS(list(
    pie_plot_data   = pie_plot_data,
    pie_base_data   = pie_base_data,
    category_levels = category_levels
  ), plot_data_rds)

  write.csv(pie_plot_data, source_data_csv, row.names = FALSE)
  cat("wrote:", plot_data_rds, "\n")
  cat("wrote:", source_data_csv, "\n")
}

print(pie_plot_data)
```

# Build the panel

```{r build_panel, fig.width=params$width, fig.height=params$height}
slice_cols <- c("G>A", "other")
custom_colors <- c("G>A" = "red2", "other" = "white")

p <- ggplot() +
  geom_scatterpie(
    data = pie_plot_data,
    aes(x = x_pos, y = y_pos, r = radius),
    cols = slice_cols,
    color = "black", linewidth = 0.7, alpha = 0.95
  ) +
  # Donut hole — white inner circle drawn in plot coordinates so it always
  # scales correctly with the pie (50% of pie radius).  v4 RMD used
  # geom_point with mm-scaled size; that broke for small pies.
  geom_circle(
    data = pie_plot_data,
    aes(x0 = x_pos, y0 = y_pos, r = radius * 0.5),
    fill = "white", colour = NA,
    inherit.aes = FALSE
  ) +
  geom_text(
    data = pie_plot_data,
    aes(x = x_pos, y = y_pos, label = format(Total, big.mark = ",")),
    fontface = "bold", size = 3.5, family = "Nimbus Sans",
    inherit.aes = FALSE
  ) +
  scale_fill_manual(values = custom_colors,
                    breaks = c("G>A", "other"),
                    labels = c("G>A (ADAR-treatable)", "other"),
                    name   = NULL) +
  scale_x_continuous(breaks = 1:3, labels = category_levels,
                     expand = expansion(mult = 0.25)) +
  scale_y_continuous(breaks = 1:3, labels = c("1", "2", "3"),
                     expand = expansion(mult = 0.25)) +
  coord_fixed() +
  labs(x = "ClinVar classification",
       y = "SFARI gene score",
       tag = "B") +    # was "C" — relabelled in composite v3 layout
  theme_publication(base_size = 12, x_angle = 0, y_bold = TRUE) +
  theme(
    legend.position    = "top",
    axis.ticks.x       = element_blank(),
    axis.ticks.y       = element_blank(),
    panel.grid.major   = element_line(color = "gray95"),
    panel.grid.minor   = element_blank(),
    plot.tag           = element_text(face = "bold", size = 16,
                                      family = "Nimbus Sans"),
    plot.tag.position  = c(0, 1)
  )

print(p)
```

# Save (PDF + ggplot RDS)

```{r save}
save_panel(p, panel_id,
           pdf_dir = out_main,
           rds_dir = intermediate_dir,
           width  = params$width,
           height = params$height)
```

# Companion summary table

Per-bucket variant counts and G>A fraction — useful for the figure caption.

```{r supp_table}
summary_table <- pie_plot_data %>%
  transmute(
    `SFARI score`   = sfari_gene_score,
    `ClinVar group` = Category,
    `G>A`           = `G>A`,
    `other`         = other,
    Total           = Total,
    `G>A pct`       = round(100 * `G>A` / pmax(Total, 1), 1)
  ) %>%
  arrange(`SFARI score`, factor(`ClinVar group`, levels = category_levels))

write.csv(summary_table,
          file.path(out_supp_tables,
                    paste0("Supp_Table_F1C_", panel_id, "_summary.csv")),
          row.names = FALSE)
print(summary_table)
```

# Session info

```{r}
sessionInfo()
```
